The state learner -- a super learner for right-censored data

### 2025-05-21 manuscript/manuscript-one.tex
divide into two manuscripts and basic template

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
\usepackage[authoryear]{natbib}
\definecolor{linkcolor}{rgb}{0, 0, 0.54}
\usepackage[colorlinks,allcolors=linkcolor]{hyperref}
%
% please place your own definitions here and don't use \def but
\DeclareMathOperator{\E}{\mathbb{E}} % expectation
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand{\weakly}{\rightsquigarrow}
\newcommand\smallO{\textit{o}}
\newcommand\bigO{\textit{O}}
\newcommand{\midd}{\; \middle|\;}
\newcommand{\1}{\mathds{1}}
\usepackage{ifthen} %% Empirical process with default argument
\newcommand{\G}[2][n]{
{\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}
\newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
\newcommand{\data}{\ensuremath{\mathcal{D}}}
%
% Insert the name of "your journal" with
\journalname{Lifetime Data Analysis}
%
\begin{document}

\title{The state learner -- a super learner for right-censored data
  % \thanks{Grants or other notes about the article that should go on the front
  % page should be placed here. General acknowledgments should be placed at the
  % end of the article.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}

\titlerunning{The state learner}        % if too long for running head

\author{Blinded author}

%\authorrunning{Short form of author list} % if too long for running head

% \institute{Anders Munch \at
%               Section of Biostatistics, University of Copenhagen, Denmark \\
%               % Tel.: +123-45-678910\\
%               % Fax: +123-45-678910\\
%               \email{a.munch@sund.ku.dk}           %  \\
% %             \emph{Present address:} of F. Author  %  if needed
%               \and %
%               Thomas~A.~Gerds %
%               \at Section of Biostatistics, University of Copenhagen, Denmark}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

@@

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone tools and in
  applications of causal inference to estimate nuisance parameters. The super
  learner is a machine learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. In right-censored
  data, the choice of the loss function and the estimation of the expected loss
  need careful consideration. We introduce the state learner, a new super
  learner for survival analysis, which simultaneously evaluates libraries of
  prediction models for the event of interest and the censoring distribution.
  The state learner can be applied to all types of survival models, works in the
  presence of competing risks, and does not require a single pre-specified
  estimator of the conditional censoring distribution. We establish an oracle
  inequality for the state learner and investigate its performance through
  numerical experiments. We illustrate the application of the state learner with
  prostate cancer data, as a stand-alone prediction tool, and, for causal
  inference, as a way to estimate the nuisance parameter models of a smooth
  statistical functional.

  \keywords{Competing risks, cross-validation, loss based estimation,
    right-censored data, super learner}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

% Medical decision making is traditionally based on the discrepancy
% between intuitive judgment and the law of chance
% \citep{redelmeier1995probability}.
% We are concerned with the statistical models that provide
% probabilistic predictions about the medical future of a patient. We
% have two different applications of these models in mind. One is as a
% stand-alone tool which predicts the likelihood of medical outcomes.
% The other is as nuisance parameter estimate
% input to an algorithm which estimates a causal target
% parameter such as an average treatment effect and depends on estimates
% of nuisance parameters such as the conditional probability of being
% uncensored.  An example of the latter is the targeted minimum loss
% based estimator \citep{rytgaard2022targeted}. In both applications,
% the models have to be learned based on available data from previous
% patients.
A super learner is a machine learning algorithm that combines a finite
set of learners into a meta learner by estimating prediction
performance in hold-out samples using a pre-specified loss function
\citep{van2007super}. When the aim is to make a prediction model,
super learners combine strong learners, such as Cox regression models
and random survival forests \citep[][Section 8.4]{gerds2021medical}.
% For targeted learning of a low dimensional
% target parameter in presence of a high-dimensional nuisance parameter,
% a super learner will usually also include the highly adaptive lasso
% \citep{van2011targeted, benkeser2016highly,van2017generally}.
While the general idea of combining strong learners based on cross-validation
stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
name super learner is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}.
% The super learner is guaranteed to perform
% almost as well as the model which minimises the expected performance, i.e., the
% model we would select if we could evaluate the prediction performance in an
% infinite hold-out sample.

We define the state learner, a new super learner for right-censored data, which
simultaneously estimates the expected loss of learners of the event time
distribution and the censoring distribution. Compared to the super learners of
\cite{polley2011-sl-cens} and \cite{golmakani2020super}, the state learner
proposed here poses no restrictions on the type of learners that can be included
in the library. Compared to super learners that rely on inverse probability of
censoring weighting, censoring unbiased transformations, or pseudo-values, the
state learner does not require a pre-specified censoring distribution. Compared
to the super learners by \cite{han2021inverse} and \cite{westling2021inference},
the state learner has theoretical guarantees which these methods do not have.
% Moreover, state learner seems to be the first super learner for
% right censored data which learns the outcome and the censoring
% distribution simulaneously.

@@

We introduce our notation and framework in Section~\ref{sec:framework}. In
Section~\ref{sec:super-learning} we define general super learning for
right-censored data. Section~\ref{sec:super-learner-simple} introduces the state
learner, and Section~\ref{sec:theor-results-prop} provides theoretical
guarantees. In Section~\ref{sec:targeted-learning} we discuss the use of the
state learner in the context of targeted learning. We report results of our
numerical experiments in Section~\ref{sec:numer-exper} and analyse a prostate
cancer data set in Section~\ref{sec:real-data-appl}.
Section~\ref{sec:discussion} contains a discussion of the merits and limitations
of our proposal. Proofs are given in Appendix~\ref{sec:proofs}. An
implementation of the state learner is available at
\url{https://github.com/amnudn/statelearner} along with code for reproducing our
numerical experiments.

@@

In survival analysis, a super learner can be used to estimate a
parameter $\Psi$ which can be identified from the observed data
distribution \(P\in\mathcal P\). In this section, to introduce the
discrete super learner and the oracle learner, we consider estimation of the
function-valued parameter \(\Psi:\mathcal P\to\mathcal M_{\tau}\), given by
\(\Psi(P)=\Lambda_j\). This parameter is identified via equation
\eqref{eq:lambdaj} on \([0,\tau]\).

As input to the super learner we need a data set
\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a collection of candidate
learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}_{\tau}\) which takes a data
set as input and returns an estimate $a(\data_n) \in \mathcal{M}_{\tau}$ of
$\Lambda_{j}$.
% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) with a loss function
\(L\colon \mathcal{M}_{\tau} \times \mathcal{O} \rightarrow \R_+\) and
estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
% The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
% learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
The oracle learner is defined as the learner that minimises the
expected loss under the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner
depend on the library of learners and on the number of folds \(K\),
and that the oracle learner is a function of the data and the unknown
data-generating distribution. However, these dependencies are
suppressed in the notation.

@@

The main idea of the state learner is to jointly use learners for
\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
distribution \( P \). The discrete state learner ranks a tuple of learners for
the tuple of the cumulative hazard functions
\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
observed data. Risk predictions can then be obtained by combining
\( \Lambda_1 \) and $\Lambda_2$ from the highest ranked tuple using a well-known
formula \citep{benichou1990estimates, ozenne2017riskregression}. To formally
introduce the state learner, we define the multi-state process

@@

for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and \( x \in \mathcal{X} \).
The function \( F \) describes the conditional state occupation probabilities of
the multi-state process \(\eta\). We construct a super learner for \( F \). The
target of this super learner is the function-valued parameter $\Psi(P) = F$
which is identified through equation~(\ref{eq:F-def}). Under conditional
independent censoring each quadruple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$
characterises a distribution \(P\in\mathcal P\), c.f.\
equation~\eqref{eq:parametrizeP}, which in turn determines \( (F, H) \). Hence,
a learner for \( F \) can be constructed from learners for \( \Lambda_1 \),
\( \Lambda_2 \), and $\Gamma$ as follows:

@@

@@

In this section we establish theoretical guarantees for the state learner. We
show that the state learner is consistent if its library contains a consistent
learner, and we establish a finite sample inequality for the excess risk of the
state learner compared to the oracle. Finally we show that a certain second
order structure is preserved when the state learner is used for targeted
learning.

@@

\begin{proposition}
  \label{prop:stric-prop}
  If \(P_0\in\mathcal{P}\) then
  \begin{equation*}
    F_0 = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]},
  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
  \end{equation*}
  \( H \)-almost surely for any \( j\in \{-1,0,1,2\} \) and almost any
  \( t \in [0, \tau]\).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:consistency-proof}.
\end{proof}

@@

\subsection{Transience of the second order remainder structure}
\label{sec:trans-second-order}

In this section we demonstrate a theoretical property of the state
learner which is useful for targeted learning (c.f., Section
\ref{sec:targeted-learning}). Specifically we consider an estimator of
a target parameter which is obtained by substituting the state learner
estimates of the nuisance parameters $\Lambda_1$, $\Lambda_2$, and
$\Gamma$. An example is an estimator of the cumulative incidence
curve, which can be obtained from estimators of $\Lambda_1$ and
$\Lambda_2$. Another example is provided in Section \ref{sec:targeted-learning}.
By equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the definition of
\( F \), we have
  \label{eq:7}
  \Gamma(t , x) 
  = \int_0^t  \frac{F(\diff s, -1, x )}{F(s-, 0, x )},
  \quad \text{and} \quad
  \Lambda_j(t , x) 
  = \int_0^t  \frac{F(\diff s, j, x )}{F(s-, 0, x )},
  \quad j \in \{1,2\},
and thus an estimator based on \( \hat{\Lambda}_{1,n} \),
\( \hat{\Lambda}_{2,n} \), and \( \hat{\Gamma}_{n} \) can also be obtained from
an estimator of $F$ using equation~(\ref{eq:7}). A so-called targeted estimator
has the key feature that it is asymptotically equivalent to a sum of i.i.d.\
random variables plus a second order remainder term
\citep{van2011targeted,hines2022demystifying}. For the setting with competing
risks, the remainder term is dominated by terms of the form
  \label{eq:dr-term}
  P{\left[
      \int_0^{\tau} w_n(s, \blank)
      \hat{M}_{1,n}(s \mid  \blank)
      \hat{M}_{2,n}(\diff s \mid  \blank)
    \right]},
where \( (\hat{M}_{1,n}, \hat{M}_{2,n}) \) is any of the nine combinations of
\( \hat{M}_{1,n} \in \{[\Gamma -\hat{\Gamma}_n], [\Lambda_1
-\hat{\Lambda}_{1,n}], [\Lambda_2 -\hat{\Lambda}_{2,n}]\} \) and
\( \hat{M}_{2,n} \in \{[\Gamma -\hat{\Gamma}_n], [\Lambda_1
-\hat{\Lambda}_{1,n}], [\Lambda_2 -\hat{\Lambda}_{2,n}]\} \), and \( w_n \) is
some data-dependent function with domain \([0,\tau]\times\mathcal X \)
\citep{van2003unified}. In particular, a targeted estimator will be
asymptotically linear if the `products' of the estimation errors
\( \hat{M}_{1,n} \) and \( \hat{M}_{2,n} \) in equation~(\ref{eq:dr-term}) are
\( \smallO_P{(n^{-1/2})}\). Proposition~\ref{prop:dr-structure} states that if
equation~(\ref{eq:dr-term}) holds for a targeted estimator based on estimators
$\hat{\Lambda}_{1,n}$, $\hat{\Lambda}_{2,n}$, and $\hat{\Gamma}_{n}$, then a
similar product structure holds for a targeted estimator based on
\( \hat{F}_n \). We state the result for the special case that
\(\hat{M}_{1,n}= \Gamma-\hat{\Gamma}_n \) and
\(\hat{M}_{2,n} =\Lambda_1-\hat{\Lambda}_{1,n} \), but similar results hold for
any combinations of \( \Gamma-\hat{\Gamma}_n\),
\( \Lambda_1-\hat{\Lambda}_{1,n} \), and \( \Lambda_2-\hat{\Lambda}_{2,n} \).
\begin{proposition}
  \label{prop:dr-structure}
  Assume that \( w(s,x)\leq c \), \( F(s, 0, x) \geq 1/c \) and
  \( \hat{F}_n(s, 0, x) \geq 1/c \) for some \( c>0 \) for all
  \( s \in [0, \tau] \) and \( x \in \mathcal{X} \). Then there are real-valued
  uniformly bounded functions \( w^a_n \), \( w^b_n \), \( w^c_n \), and
  \( w^d_n \) with domain \( [0,\tau]^2 \times \mathcal{X} \) such that
    & P{\left[
      \int_0^{\tau} w(s, \blank)
      \left\{
      \Gamma(s,\blank) -\hat{\Gamma}_n(s,\blank)
      \right\}
      [\Lambda_1-\hat{\Lambda}_{1,n}]
      (\diff s, \blank)
      \right]}
      P{\big[
      \int_0^{\tau} \int_0^{s} w^a_n(s,u,\blank) [F - \hat{F}_n](u-, 0,
       \blank)[F - \hat{F}_n](s-, 0, \blank)}
    & \qquad \quad \times F(\diff u, -1, \blank ) F ( \diff s, 1, \blank)
      \big]
    & \quad +
      P{\left[
      \int_0^{\tau} \int_0^{s} w^b_n(s,u,\blank) [F - \hat{F}_n](u-, 0, \blank)

>>>>

\documentclass[a4paper,danish]{article}
\usepackage[margin=3.5cm]{geometry}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{natbib, dsfont, enumitem,
  amssymb,soul,xcolor,amsmath,amsthm,graphicx,subcaption,verbatim,pgfplots,tikz,prodint}
\usetikzlibrary{calc,patterns,angles,quotes,automata,
  positioning,arrows,shapes}

\definecolor{linkcolor}{rgb}{0, 0, 0.54}
\usepackage[colorlinks,allcolors=linkcolor]{hyperref}

% Ref and bibliography style 
\bibliographystyle{abbrvnat}

% Handling new lines
\setlength{\parskip}{1em}
\setlength{\parindent}{0em}

% Handling space after sections
\usepackage{titlesec}
\titlespacing*{\section}{0em}{2em}{0em}
\titlespacing*{\subsection}{0em}{2em}{0em}
\titlespacing*{\subsubsection}{0em}{2em}{0em}

% No spacing after in start of list
\setlist[itemize]{topsep=0pt}
\setlist[enumerate]{topsep=0pt}

% Theorem environments 
\theoremstyle{plain} % plain italic style
\newtheorem{theorem}{Theorem}
\numberwithin{theorem}{section}
\newtheorem*{theorem*}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition} % non-italic
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem*{assumption*}{Assumption}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem{note}[theorem]{Note}

% Notation
\DeclareMathOperator{\E}{\mathbb{E}} % expectation
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand{\weakly}{\rightsquigarrow}
\newcommand\smallO{\textit{o}}
\newcommand\bigO{\textit{O}}
\newcommand{\midd}{\; \middle|\;}
\newcommand{\1}{\mathds{1}}
\usepackage{ifthen} %% Empirical process with default argument
\newcommand{\G}[2][n]{
{\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}
\newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
\newcommand{\data}{\ensuremath{\mathcal{D}}}

\title{The state learner -- a super learner for right-censored
  data}
\author{Anders Munch \& Thomas A. Gerds}
\date{\today}

\begin{document}
\maketitle

% %%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
% %
% % This is a general template file for the LaTeX package SVJour3
% % for Springer journals.          Springer Heidelberg 2010/09/16
% %
% % Copy it to a new file with a new name and use it as the basis
% % for your article. Delete % signs as needed.
% %
% % This template includes a few options for different layouts and
% % content for various journals. Please consult a previous issue of
% % your journal as needed.
% %
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %
% % First comes an example EPS file -- just ignore it and
% % proceed on the \documentclass line
% % your LaTeX will extract the file if required
% \begin{filecontents*}{example.eps}
% %!PS-Adobe-3.0 EPSF-3.0
% %%BoundingBox: 19 19 221 221
% %%CreationDate: Mon Sep 29 1997
% %%Creator: programmed by hand (JK)
% %%EndComments
% gsave
% newpath
%   20 20 moveto
%   20 220 lineto
%   220 220 lineto
%   220 20 lineto
% closepath
% 2 setlinewidth
% gsave
%   .4 setgray fill
% grestore
% stroke
% grestore
% \end{filecontents*}
% %
% \RequirePackage{fix-cm}
% %
% %\documentclass{svjour3}                     % onecolumn (standard format)
% %\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
% \documentclass[smallextended]{svjour3}       % onecolumn (second format)
% %\documentclass[twocolumn]{svjour3}          % twocolumn
% %
% \smartqed  % flush right qed marks, e.g. at end of proof
% %
% \usepackage{graphicx}
% %
% % \usepackage{mathptmx}      % use Times fonts if available on your TeX system
% %
% % insert here the call for the packages your document requires
% \usepackage{dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
% \usepackage[authoryear]{natbib}
% \definecolor{linkcolor}{rgb}{0, 0, 0.54}
% \usepackage[colorlinks,allcolors=linkcolor]{hyperref}
% %
% % please place your own definitions here and don't use \def but
% \DeclareMathOperator{\E}{\mathbb{E}} % expectation
% \newcommand{\Z}{\mathbb{Z}}
% \newcommand{\R}{\mathbb{R}}
% \newcommand{\N}{\mathbb{N}}
% \newcommand{\C}{\mathbb{C}}
% \renewcommand{\S}{\mathbb{S}}
% \newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
% \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
% \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
% \renewcommand{\phi}{\varphi}
% \renewcommand{\epsilon}{\varepsilon}
% \newcommand*\diff{\mathop{}\!\mathrm{d}}
% \newcommand{\weakly}{\rightsquigarrow}
% \newcommand\smallO{\textit{o}}
% \newcommand\bigO{\textit{O}}
% \newcommand{\midd}{\; \middle|\;}
% \newcommand{\1}{\mathds{1}}
% \usepackage{ifthen} %% Empirical process with default argument
% \newcommand{\G}[2][n]{
% {\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
% }
% \DeclareMathOperator*{\argmin}{\arg\!\min}
% \DeclareMathOperator*{\argmax}{\arg\!\max}
% \newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
% \newcommand{\data}{\ensuremath{\mathcal{D}}}

% \begin{document}

% \title{The state learner -- a super learner for right-censored
%   data}

% \author{Anders Munch \& Thomas A. Gerds}

@@

\begin{abstract}
  In survival analysis, prediction models are needed as
  stand-alone tools and in applications of causal inference to
  estimate nuisance parameters. The super learner is a machine
  learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. In
  right-censored data, the choice of the loss function and the
  estimation of the expected loss need careful consideration.
  We introduce the state learner, a new super learner for
  survival analysis, which simultaneously evaluates libraries
  of prediction models for the event of interest and the
  censoring distribution. The state learner places no
  restrictions on the algorithms that can be included in the
  library, works in the presence of competing risks, and does
  not require a single pre-specified estimator of the
  conditional censoring distribution. We establish an oracle
  inequality for the state learner and investigate its
  performance through numerical experiments. We illustrate the
  application of the state learner with prostate cancer data.

  % \keywords{Competing risks, cross-validation, loss based estimation,
  %   right-censored data, super learner}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

[More motivation]

A super learner is a machine learning algorithm that combines
a finite set of learners into a meta learner by estimating
prediction performance in hold-out samples using a
pre-specified loss function \citep{van2007super}. When the aim
is to make a prediction model, super learners combine strong
learners, such as Cox regression models and random survival
forests \citep[][Section 8.4]{gerds2021medical}. While the
general idea of combining strong learners based on
cross-validation stems from earlier work
\citep{wolpert1992stacked,breiman1996stacked}, the name super
learner is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}.

We define the state learner, a new super learner for
right-censored data, which simultaneously estimates the
expected loss of learners of the event time distribution and
the censoring distribution. Compared to the super learners of
\cite{polley2011-sl-cens} and \cite{golmakani2020super}, the
state learner proposed here poses no restrictions on the type
of learners that can be included in the library. Compared to
super learners that rely on inverse probability of censoring
weighting, censoring unbiased transformations, or
pseudo-values, the state learner does not require a
pre-specified censoring distribution. Compared to the super
learners by \cite{han2021inverse} and
\cite{westling2021inference}, the state learner has
theoretical guarantees which these methods do not have.

@@

We introduce our notation and framework in
Section~\ref{sec:framework}. In
Section~\ref{sec:super-learning} we define general super
learning for right-censored data.
Section~\ref{sec:super-learner-simple} introduces the state
learner, and Section~\ref{sec:theor-results-prop} provides
theoretical guarantees. We report results of our numerical
experiments in Section~\ref{sec:numer-exper} and analyse a
prostate cancer data set in Section~\ref{sec:real-data-appl}.
Section~\ref{sec:discussion} contains a discussion of the
merits and limitations of our proposal. Proofs are given in
Appendix~\ref{sec:proofs}. An implementation of the state
learner is available at
\url{https://github.com/amnudn/statelearner} along with code
for reproducing our numerical experiments.

@@

In survival analysis, a super learner can be used to estimate
a parameter $\theta \in \Theta$ which can be identified from
the observed data distribution \(P\in\mathcal P\). A super
learner typically estimates a function, and thus the parameter
space $\Theta$ is most often a class of functions. For
example, $\theta$ could be a conditional survival function or
a cumulative hazard function. In
Section~\ref{sec:super-learner-simple} we consider a specific
choice of parameter as target for our super learner, but in
this section we describe super learning for a general
parameter $\theta$ with values in some parameter space
$\Theta$.

As input to a super learner we need a data set
\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from
\( P \in \mathcal{P} \) and a collection of candidate learners
$\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
\( a \colon \mathcal{O}^n \rightarrow \Theta \) which takes a
data set as input and returns an estimate
$a(\data_n) \in \Theta$ of $\theta$.
% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) with a loss function
\(L\colon \Theta \times \mathcal{O} \rightarrow \R_+\) and
estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
The oracle learner is defined as the learner that minimises the
expected loss under the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner
depend on the library of learners and on the number of folds \(K\),
and that the oracle learner is a function of the data and the unknown
data-generating distribution. However, these dependencies are
suppressed in the notation.

@@

The main idea of the state learner is to jointly use learners
for \( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and
the relations in equation~(\ref{eq:parametrizeP}), to learn a
feature of the observed data distribution \( P \). The
discrete state learner ranks a tuple of learners for the tuple
of the cumulative hazard functions
\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they
jointly model the observed data. To formally introduce the
state learner, we define the multi-state process

@@

for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and
\( x \in \mathcal{X} \). The function \( F \) describes the
conditional state occupation probabilities of the multi-state
process \(\eta\). We construct a super learner for \( F \).
The target of this super learner is the function-valued
parameter $\theta(P) = F$ which is identified through
equation~(\ref{eq:F-def}). Under conditional independent
censoring each quadruple
$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a
distribution \(P\in\mathcal P\), c.f.\
equation~\eqref{eq:parametrizeP}, which in turn determines
\( (F, H) \). Hence, a learner for \( F \) can be constructed
from learners for \( \Lambda_1 \), \( \Lambda_2 \), and
$\Gamma$ as follows:

@@

\section{Use cases for the state learner}
\label{sec:use-cases-state}

The parameter \( F \) is not of much interest...

\subsection{Survival and risk predictions}
\label{sec:surv-risk-pred}

...

Risk predictions can then be obtained by combining
\( \Lambda_1 \) and $\Lambda_2$ from the highest ranked tuple
using a well-known formula \citep{benichou1990estimates,
  ozenne2017riskregression}.

\subsection{Targeted and debiased machine learning}
\label{sec:targeted-learning}

... TL + SL = strong... The state learner is particularly well
suited for targeted and debiased machine learning as it
immediately provides estimates of the nuisance parameters
pertaining to the outcome and the censoring models. While it
was our original motivation for developing the state learning,
we leave the study of the state learner in the context of
targeted and debiased machine learning for a future paper.

@@

In this section we establish theoretical guarantees for the
state learner. We show that the state learner is consistent if
its library contains a consistent learner, and we establish a
finite sample inequality for the excess risk of the state
learner compared to the oracle.

@@

\begin{proposition}
  \label{prop:stric-prop}
  If \(P_0\in\mathcal{P}\) then
  \begin{equation*}
    F_0 = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]},
  \end{equation*}
  \( H \)-almost surely for any \( j\in \{-1,0,1,2\} \) and almost any
  \( t \in [0, \tau]\).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:consistency-proof}.
\end{proof}

@@
